---
title: "Analysis 1"
author: "James Ashmore"
date: "2019-10-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Setup

Set knitr chunk options:

```{r knitr}
knitr::opts_chunk$set(
  autodep    = TRUE,
  cache      = TRUE,
  cache.path = "cache/01-analysis.Rmd/",
  dev        = "png",
  error      = FALSE,
  message    = FALSE,
  warning    = FALSE
)
```

Load required analysis packages:

```{r}
pacman::p_load(scater, scran, SingleCellExperiment, uwot, Rtsne)
```

Import experiment data:

```{r}
sce <- readr::read_rds(here::here("data/annotation.rds"))
```

## Exploration

Ectoderm, paraxial and lateral mesoderm markers:

```{r}
ids <- data.frame(
  group = factor(c(1, 1, 1, 1, 2, 2, 2, 2, 3, 3, 4, 4, 4, 4, 5, 5, 5, 5, 5, 6, 6, 6, 6, 6)),
  row.names = c("Vim", "Twist2", "Twist1", "Snai1", "Runx1", "Tal1", "Etv2", "Kdr", "Hoxa11", "Hoxa10", "Aldh1a2", "Pitx1", "Foxf1", "Hand1", "Msgn1", "Pcdh19", "Tbx6", "Dll3", "Dll1", "Cldn7", "Tfap2c", "Nkx1-2", "Cdh1", "Epcam")
)
```

Plot scaled expression values of markers:

```{r fig.width = 8}
mat <- logcounts(sce)

mat <- mat[rownames(ids), ]

mat <- mat[order(ids$group), ]

mat <- mat[, order(sce$cluster)]

mat <- mat - rowMeans(mat)

val <- max(abs(mat))

lim <- c(-val, val)

mat[mat < lim[1]] <- lim[1]

mat[mat > lim[2]] <- lim[2]

pal <- colorRampPalette(rev(RColorBrewer::brewer.pal(5, name = "RdYlBu")))

col <- pal(100)

brk <- seq(lim[1], lim[2], length.out = 101)

ann <- list(
  row = data.frame(Markers = ids$group, row.names = rownames(ids)),
  col = data.frame(Cluster = sce$cluster, row.names = colnames(sce))
)

pheatmap::pheatmap(
  mat, 
  color = col, 
  breaks = brk, 
  scale = "none", 
  cluster_rows = FALSE, 
  cluster_cols = FALSE, 
  annotation_row = ann$row, 
  annotation_col = ann$col,
  show_colnames = TRUE
)
```

Streak markers:

```{r}
ids <- data.frame(
  group = factor(c(1, 1, 1, 1, 1, 2, 2, 2, 3, 3, 3, 3, 3, 4, 4, 5, 5, 6, 6, 7, 7, 8, 8)),
  row.names = c("Hoxa5", "Hoxa11", "Hoxa10", "Hoxa9", "Hoxa7", "Sox2", "Eomes", "Nanog", "Sp6", "Wnt8a", "Wnt3a", "Fgf10", "Fgf8", "Snai1", "Vim", "Etv2", "Kdr", "Foxf1", "Hand1", "Tbx6", "Dll3", "Cdh1", "Epcam")
)
```

Plot scaled expression values of markers:

```{r fig.width = 8}
mat <- logcounts(sce)

mat <- mat[rownames(ids), ]

mat <- mat[order(ids$group), ]

mat <- mat[, order(sce$cluster)]

mat <- mat - rowMeans(mat)

val <- max(abs(mat))

lim <- c(-val, val)

mat[mat < lim[1]] <- lim[1]

mat[mat > lim[2]] <- lim[2]

pal <- colorRampPalette(rev(RColorBrewer::brewer.pal(5, name = "RdYlBu")))

col <- pal(100)

brk <- seq(lim[1], lim[2], length.out = 101)

ann <- list(
  row = data.frame(Markers = ids$group, row.names = rownames(ids)),
  col = data.frame(Cluster = sce$cluster, row.names = colnames(sce))
)

pheatmap::pheatmap(
  mat, 
  color = col, 
  breaks = brk, 
  scale = "none", 
  cluster_rows = FALSE, 
  cluster_cols = FALSE, 
  annotation_row = ann$row, 
  annotation_col = ann$col,
  show_colnames = FALSE
)
```

Manually annotate the clusters using information from Valerie Wilson:

```{r}
tbl <- read.csv(here::here("data/cell_annotation.csv"))

sce$germ_layer <- tbl$germ_layer[match(sce$Sample, tbl$cell_id)]
```

Ectoderm, paraxial and lateral mesoderm and streak markers:

```{r}
ids <- c("Vim", "Twist2", "Twist1", "Snai1", "Runx1", "Tal1", "Etv2", "Kdr", "Hoxa11", "Hoxa10", "Aldh1a2", "Pitx1", "Foxf1", "Hand1", "Msgn1", "Pcdh19", "Tbx6", "Dll3", "Dll1", "Cldn7", "Tfap2c", "Nkx1-2", "Cdh1", "Epcam", "Hoxa5", "Hoxa9", "Hoxa7", "Sox2", "Eomes", "Nanog", "Sp6", "Wnt8a", "Wnt3a", "Fgf10", "Fgf8")
```

Plot scaled expression values of markers:

```{r fig.width = 8}
mat <- logcounts(sce)

mat <- mat[ids, ]

mat <- mat[, order(sce$germ_layer)]

mat <- mat - rowMeans(mat)

val <- max(abs(mat))

lim <- c(-val, val)

mat[mat < lim[1]] <- lim[1]

mat[mat > lim[2]] <- lim[2]

pal <- colorRampPalette(rev(RColorBrewer::brewer.pal(5, name = "RdYlBu")))

col <- pal(100)

brk <- seq(lim[1], lim[2], length.out = 101)

ann <- list(
  col = data.frame(
    Cluster = sce$cluster,
    Layer = sce$germ_layer,
    row.names = colnames(sce)
  )
)

pheatmap::pheatmap(
  mat, 
  color = col, 
  breaks = brk, 
  scale = "none", 
  cluster_rows = TRUE, 
  cluster_cols = FALSE, 
  annotation_col = ann$col,
  show_colnames = FALSE
)
```

## Summary

Save experiment object to disk:

```{r}
saveRDS(sce, here::here("data/01-analysis.rds"))
```
